Staged Abstract Interpreters

### 2019-09-06 paper/50_stageabs.tex
update

Every components in the interpreter except @Expr@s are staged. The data
structures representing these components are treated as black-boxes, such as
@Map@, which means that the operations on a @Map@ directly become nex-stage code,
but we don't inspect any further inside. As we identified when
introducing the generic interface, the keys of @Env@s are identifiers (i.e.,
strings) in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures.  For
example, let us assume that the @Map[K, V]@ is implemented as a hash map. If
all the keys of type @K@ are all known statically, then the indices also can be
computed statically. Thus, the specialized map would be an array of type
@Array[Rep[V]]@, whose elements are next-stage values, and the size of the array is
known statically. An access to the map is translated into an access to the
array with the pre-computed index during staging.

@@

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive calls in our abstract interpreter. But this
is not necessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in A-Normal Form
$\lambda$-calculus \cite{Flanagan:1993:ECC:155090.155113}, we can use a
selective caching algorithm that does not generate caching code for atomic
expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

@@

We implemented a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive calls of our abstract
interpreter is applied to the proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of the specialization procedure. It is also possible to specialize a small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} presented for
optimizing Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering effort.

>>>>

All components in the interpreter except @Expr@s are staged. The data
structures representing these components are treated as black-boxes, such as
@Map@, which means that the operations on a @Map@ directly become nex-stage code,
but we don't inspect any further inside. As we identified when
introducing the generic interface, the keys of @Env@s are identifiers (i.e.,
strings) in the program, which are completely known statically. This
observation gives us a chance to further specialize the data structures.  For
example, let us assume that the @Map[K, V]@ is implemented as a hash map. If
all the keys of type @K@ are all known statically, then the indices also can be
computed statically. Thus, the specialized map would be an array of type
@Array[Rep[V]]@, whose elements are next-stage values, and the size of the array is
known statically. An access to the map is translated into an access to the
array with the pre-computed index during staging.

@@

\paragraph{Selective Caching} We observe that the two-fold co-inductive
caching is used for every recursive call in our abstract interpreter. But this
is not necessary and redundant when generating code for atomic
expressions such as literals or variables, because they always terminate.
Borrowing the idea from the partition of expressions used in A-Normal Form
$\lambda$-calculus \cite{Flanagan:1993:ECC:155090.155113}, we can use a
selective caching algorithm that does not generate caching code for atomic
expressions:
\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

@@

We implemented a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to the proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of the specialization procedure. It is also possible to specialize a small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} presented for
optimizing Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering effort.

### 2019-09-06 paper/60_cases.tex
revision
update

We first revisit a similar technique called \textit{abstract compilation} (AC).
AC was introduced by \citet{Boucher:1996:ACN:647473.727587} as an
implementation technique for abstract interpretation-based static analysis.
Similar to the present paper, the idea is inspired by partial evaluation:
the program is known statically and the interpretive overhead can be
eliminated. In AC, the compiled analysis can be represented by either text or
closures (higher-order functions). The minute difference is that the closures
can be executed immediately, while the textual programs need to be compiled and
loaded.

Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} showed how to compile a
monovariant control-flow analysis \cite{Shivers:1991:SSC:115865.115884,
Shivers:1988:CFA:53990.54007} for continuation-passing style (CPS) programs.
Since the analyzed program is in CPS form, the analyzer is a big-step
control-environment abstract interpreter. The abstract compilation transforms
the analysis as a closure just taking an environment argument. Therefore the
overhead of traversing the AST of input program has been eliminated.

In this section, we show that the AC (closure generation as example) can be
understood and implemented as an instance of staging abstract interpreters. We
first revisit the original implementation of abstract compilation of 0-CFA, and
then reproduce their result by simply adding stage annotations.
The program generated by our approach provides approximately the same improvement with regards to
speed, but without changing a single line of the analyzer program using
type-based staging annotations. However, AC requires more engineering effort,
i.e., rewriting the whole analyzer into the closure-generation form. Moreover,
as shown in Section~\ref{staged_ds}, our approach is able to not only remove the
interpretive overhead, but also enable several optimizations such as specialized
data structures and equivalent rewritings.

@@

We use Scheme as an example, and consider there is a library that contains a
list map function @map@:
\begin{lstlisting}
  (define (map xs f) (if (null? xs) '() (cons (f (car xs)) (map (cdr xs) f))))
\end{lstlisting}

>>>>

We first revisit a similar technique called \textit{abstract compilation} (AC).
AC was introduced by \citet{Boucher:1996:ACN:647473.727587} as an
implementation technique for abstract interpretation-based static analyses.
Similar to the present paper, the idea of AC was inspired by partial evaluation:
the program is known statically and the interpretive overhead can be
eliminated. In AC, the compiled analysis can be represented by either text or
closures (higher-order functions). The minute difference is that the closures
can be executed immediately, while the textual programs need to be compiled and
loaded.

Specifically, \citeauthor{Boucher:1996:ACN:647473.727587} showed how to compile a
monovariant control-flow analysis \cite{Shivers:1991:SSC:115865.115884,
Shivers:1988:CFA:53990.54007} for continuation-passing style (CPS) programs.
The analyzer is a big-step control-environment abstract interpreter. 
By applying AC, we refactor every function in the analyzer such that it
return a closure just taking an environment argument.  Therefore the overhead
of traversing the AST of input program is eliminated.

In this section, we show AC (closure generation) can be understood and
implemented as an instance of staging abstract interpreters. We first revisit
the original implementation of abstract compilation for 0-CFA, and then
reproduce their result by simply adding stage annotations.
The program generated by our approach provides approximately the same
improvement with regards to speed, but without changing a single line of the
analyzer program using type-based staging annotations. However, AC requires
more engineering effort, i.e., rewriting the whole analyzer into the
closure-generation form. Moreover, as shown in Section~\ref{staged_ds}, our
approach is able to not only remove the interpretive overhead, but also enable
several optimizations such as specialized data structures and equivalent
rewritings.

@@

We use Scheme as an example, and consider a library that contains a
function @map@:
\begin{lstlisting}
  (define (map xs f) (if (null? xs) '() (cons (f (car xs)) (map (cdr xs) f))))
\end{lstlisting}
